---
title: "Bookdown Demo"
author: "Ulrike Grömping"
date: "23 June 2020"
encoding: utf8
output: 
  bookdown::html_document2:
    toc: true
    toc_float: yes
    code_folding: hide
  bookdown::pdf_document2:
    toc: false
    keep_tex: true
---

(ref:latexInCaption) A caption can contain _underscores_ and $\sqrt v$ formulae.

```{r setup, include=FALSE}
params <- list(html=knitr::is_html_output())
knitr::opts_chunk$set(echo = params$html)
```

# First chapter

The R package **bookdown** is documented in an online book by its author Yihui Xie (2020). In this brief demo, I demonstrate some basic aspects of the use of the R package **bookdown** that took my some more time to figure out from that documentation.

In Chapter&nbsp;\@ref(sec:meinreflabel2) you find a figure with a simple caption that was created by the `fig.cap` entry in the chunk information, and a table that was created with function `knitr::kable`. In Chapter&nbsp;\ref{sec:meinreflabel3} you find a figure whose caption contains underscores and a formula; because of these complications, the figure caption has to be defined as a reference (see above, "(ref:latexInCaption)") before it can be used in the chunk information; I have made it a personal convention to always define all such references at the top of the file. Such a reference can be used everywhere in the text. If you have more than one reference definition, always have an empty line between any two such definitions (there is only one here).

You can have unnumbered chapters (while most chapters are numbered), see the appendix.

With the **bookdown** package, it is easy to reference back an forth between sections, and to figures or tables that were created from code chunks. References to figures start with "fig:", references to tables with "tab:", and section references are entirely defined in the curly brackets after the heading, starting with "#".  The syntax for **bookdown** references is "\@ref(your reference target)",  see examples in this file. They are translated to the appropriate way of referencing for pdf or html files, as appropriate.

# Second chapter{#sec:meinreflabel2}

Table&nbsp;\@ref(tab:cor) shows the correlations in the dataset `swiss`.

```{r cor}
knitr::kable(round(cor(swiss),4), caption="The correlation matrix of the `swiss` data", booktabs=TRUE, linesep="")
```

Figure&nbsp;\@ref(fig:kde) shows a kernel density estimate of the proportion of males involved in agricultural occupations.

```{r kde, fig.cap="Proportion of males involved in agricultural occupations in 47 French speaking swiss provinces (1888). Kernel density estimate.", fig.height=3, fig.width=4}
plot(density(swiss$Agriculture), xlab="Agriculture")
```

# Third chapter{#sec:meinreflabel3}

Figure&nbsp;\@ref(fig:scat) shows a scatter plot with a more demanding figure caption.

```{r scat, fig.cap="(ref:latexInCaption)", fig.height=3, fig.width=4}
plot(Fertility~Agriculture, swiss)
```

# Final chapter
Your final chapter \textendash&nbsp;like your introductury chapter \textendash&nbsp;will be read by every reader. Both these chapters are therefore particularly important. 

Please note that this brief text focused on how to implement a report with the help of R package **bookdown**. There was not content of specific relevance for data visualization.

# References{-}
Xie, Y. (2020). bookdown: Authoring Books and Technical Documents with R Mardown. https://bookdown.org/yihui/bookdown/ (Status: June 17, 2020).

# Appendix A{-}
Example for an appendix

The following listing shows the data (do not do that with a large data set!).

```{r printdat}
print(swiss)
```